Stage-Splitting a Modal Language

### 2014-10-01 paper/example.tex

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in \ref{fig:quickselect}.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

>>>>

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in \ref{fig:quickselect}.
Quickselect is an algorithm for finding the $k$-th largest element in a list.
It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).

The benefits are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.
